Fluent statistical computing interfaces for biological data analysis

### 2020-10-28 Rmd/06-conclusion.Rmd
capitalisation

In this thesis, I have designed tools to explore workflow steps
that are integral to modern biological data science. In particular, I have
implemented software that facilitates the wrangling, integration, and 
visualisation of high-throughput biological data in a principled and pragmatic
manner.  The early chapters of this thesis explored the *tidy data* semantic and 
its extension to range based genomics data. This culminated in the development
of "plyranges: a grammar of genomic data transformation" in chapter 
\@ref(ch:plyranges), which developed a new domain specific language for genomics
data analysis. The applicability of the **plyranges** interface and use of the 
*tidy data* concept were further interrogated in chapter \@ref(ch:fluentGenomics),
"Fluent genomics with plyranges and tximeta", which described techniques 
integrating data along the genome, and emphasised the importance of 
interoperability between analysis tools. Similarly, chapter  \@ref(ch:intron), 
"Exploratory coverage analysis with superintronic and 
plyranges", tackled data integration from a different angle by looking at 
multiple summaries of variables measured along the genome to find putative 
regions of intron retention. In the final part of the thesis I moved towards 
visualisation issues as they related to working with high-dimensional data
common in biological data science. Chapter \@ref(ch:tsne), "Casting multiple shadows:  high-dimensional interactive data visualisation with tours and embeddings",  explored pragmatic approaches to high dimensional data 
visualisation in light of the rise of popular non-linear embedding methods.

@@

The **plyranges** package develops a suite of verbs for interacting
with genomic data as a *GRanges* object. Since its release on Bioconductor,
it  has been relatively  successful: it has been downloaded `r nice(plyranges_stats$no_downloads)` times from  `r nice(plyranges_stats$no_ip)` unique IP addresses. I have also had the privilege of teaching workshops on **plyranges**
at Bioconductor conferences which also led to the development of the 
**fluentGenomics** workflow package, outlined in chapter \@ref(ch:fluentGenomics). A broader impact of the work, has been the discussions around the
concepts of fluent interfaces and tidy data within the Bioconductor community,
which has led to several developments currently in place that are exploring
different approaches for fluent interfaces for other types of omics data.
The **plyranges** package is available to download from 
https://bioconductor.org/packages/plyranges and
the **fluentGenomics** workflow is available to download from 
https://bioconductor.org/packages/release/workflows/html/fluentGenomics.html.

The **superintronic** software described in chapter \@ref(ch:intron) has
been used in @Lee2019-mf to disentangle and view intron signal in RNA-seq
data. Here, we again show the strengths of providing a long-form representations
of genomics data (in this case coverage vectors). By leveraging **plyranges**
we were then able to create a set of data descriptors that we could link back
to the raw data to discover genes thought to be associated with a real biological 
signal. An interesting extension to this work would be applying it to single
cell and long-read based transcriptomics data, where scalability and much
larger design matrices would become an issue. The **superintronic** package
is available to download from https://github.com/sa-lee/superintronic.

@@

We showed in chapter \@ref(ch:tsne) that tours provide a global overview
that can be used as tool for exploring model fits. An issue that arises is
how to scale the tour as the number of observations increases. There are
latencies in sending data from the back end to the visualisation client
that causes lag during animation. One could also question whether point
based displays are appropriate in this case, and it would be worth exploring 
the usability of animations based on binning the projections. Moreover, 
when the number of observations are large, the points in the projections are 
concentrated in the centre of the tour display obscuring interesting aspects
of the data. This is mitigated via having
the ability to zoom, but further research into transforming the projections
to avoid crowding would be valuable. An added complexity to changes in visual
displays are thinking about the design of user interactions, and several promising
avenues based on section tours could be explored [@Laa2020-wr; @Laa2020-fy].

>>>>

In this thesis, I have designed tools to explore workflow steps
that are integral to modern biological data science. In particular, I have
implemented software that facilitates the wrangling, integration, and 
visualisation of high-throughput biological data in a principled and pragmatic
manner.  The early chapters of this thesis explored the *tidy data* semantic and 
its extension to range based genomics data. This culminated in the development
of "plyranges: a grammar of genomic data transformation" in Chapter 
\@ref(ch:plyranges), which developed a new domain specific language for genomics
data analysis. The applicability of the **plyranges** interface and use of the 
*tidy data* concept were further interrogated in Chapter \@ref(ch:fluentGenomics),
"Fluent genomics with plyranges and tximeta", which described techniques 
integrating data along the genome, and emphasised the importance of 
interoperability between analysis tools. Similarly, Chapter  \@ref(ch:intron), 
"Exploratory coverage analysis with superintronic and 
plyranges", tackled data integration from a different angle by looking at 
multiple summaries of variables measured along the genome to find putative 
regions of intron retention. In the final part of the thesis I moved towards 
visualisation issues as they related to working with high-dimensional data
common in biological data science. Chapter \@ref(ch:tsne), "Casting multiple shadows:  high-dimensional interactive data visualisation with tours and embeddings",  explored pragmatic approaches to high dimensional data 
visualisation in light of the rise of popular non-linear embedding methods.

@@

The **plyranges** package develops a suite of verbs for interacting
with genomic data as a *GRanges* object. Since its release on Bioconductor,
it  has been relatively  successful: it has been downloaded `r nice(plyranges_stats$no_downloads)` times from  `r nice(plyranges_stats$no_ip)` unique IP addresses. I have also had the privilege of teaching workshops on **plyranges**
at Bioconductor conferences which also led to the development of the 
**fluentGenomics** workflow package, outlined in Chapter \@ref(ch:fluentGenomics). A broader impact of the work, has been the discussions around the
concepts of fluent interfaces and tidy data within the Bioconductor community,
which has led to several developments currently in place that are exploring
different approaches for fluent interfaces for other types of omics data.
The **plyranges** package is available to download from 
https://bioconductor.org/packages/plyranges and
the **fluentGenomics** workflow is available to download from 
https://bioconductor.org/packages/release/workflows/html/fluentGenomics.html.

The **superintronic** software described in Chapter \@ref(ch:intron) has
been used in @Lee2019-mf to disentangle and view intron signal in RNA-seq
data. Here, we again show the strengths of providing a long-form representations
of genomics data (in this case coverage vectors). By leveraging **plyranges**
we were then able to create a set of data descriptors that we could link back
to the raw data to discover genes thought to be associated with a real biological 
signal. An interesting extension to this work would be applying it to single
cell and long-read based transcriptomics data, where scalability and much
larger design matrices would become an issue. The **superintronic** package
is available to download from https://github.com/sa-lee/superintronic.

@@

We showed in Chapter \@ref(ch:tsne) that tours provide a global overview
that can be used as tool for exploring model fits. An issue that arises is
how to scale the tour as the number of observations increases. There are
latencies in sending data from the back end to the visualisation client
that causes lag during animation. One could also question whether point
based displays are appropriate in this case, and it would be worth exploring 
the usability of animations based on binning the projections. Moreover, 
when the number of observations are large, the points in the projections are 
concentrated in the centre of the tour display obscuring interesting aspects
of the data. This is mitigated via having
the ability to zoom, but further research into transforming the projections
to avoid crowding would be valuable. An added complexity to changes in visual
displays are thinking about the design of user interactions, and several promising
avenues based on section tours could be explored [@Laa2020-wr; @Laa2020-fy].

### 2020-10-28 Rmd/07-appendix-A.Rmd
capitalisation

@@

## Getting started with the **plyranges** package

### `Ranges` revisited

In Bioconductor there are two classes, `IRanges` and `GRanges`, that are
standard data structures for representing genomics data.  Throughout this
document I refer to either of these classes as `Ranges` if an operation can be
performed on either class, otherwise I explicitly mention if a function is
appropriate for an `IRanges` or `GRanges`.

`Ranges` objects can either represent sets of integers as `IRanges` (which have
start, end and width attributes) or represent genomic intervals (which have
additional attributes, sequence name, and strand) as `GRanges`. In addition,
both types of `Ranges` can store information about their intervals as metadata
columns (for example GC content over a genomic interval).

`Ranges` objects follow the tidy data principle: each row of a `Ranges` object
corresponds to an interval, while each column will represent a variable about
that interval, and generally each object will represent a single unit of
observation (like gene annotations).

Consequently, `Ranges` objects provide a powerful representation for reasoning
about genomic data. In this vignette, you will learn more about `Ranges`
objects and how via grouping, restriction and aggregation you can perform
common data tasks.

### Constructing `Ranges`

To construct an `IRanges` we require that there are at least two columns that
represent at either a starting coordinate, finishing coordinate or the width of
the interval.

@@

To construct a `GRanges` we require a column that represents that sequence name
( contig or chromosome id), and an optional column to represent the
strandedness of an interval.

@@

### Arithmetic on Ranges

@@

```{r anchor_fig, echo = FALSE, out.width="400px", fig.align="center"}
knitr::include_graphics(here::here("img", "anchors.png"), dpi = 150)
```

@@

`Ranges` can be shifted left or right. If strand information is available we
can also shift upstream or downstream.

@@

### Grouping `Ranges`

**plyranges** introduces a new class of `Ranges` called `RangesGrouped`, this is
a similar idea to the grouped `data.frame\tibble` in **dplyr**.

Grouping can act on either the core components or the metadata columns of a
`Ranges` object.

@@

### Restricting `Ranges`

The verb `filter` can be used to restrict rows in the `Ranges`. Note that
grouping will cause the `filter` to act within each group of the data.
```{r}
grng %>% filter(gc < 0.3)
# filtering by group
grng_by_strand %>% filter(gc == max(gc))
```

We also provide the convenience methods `filter_by_overlaps` and
`filter_by_non_overlaps` for restricting by any overlapping `Ranges`.

@@

### Summarising `Ranges`

The `summarise` function will return a `DataFrame` because the information
required to return a `Ranges` object is lost. It is often most useful to use
`summarise()` in combination with the `group_by()` family of functions.

@@

### Joins, or another way at looking at overlaps between `Ranges`

@@

```{r olaps-diagram, echo = FALSE, out.width="600px"}
knitr::include_graphics(here::here("img", "olap-fig.png"))
```

The first function, `join_overlap_intersect()` will return a `Ranges` object
where the start, end, and width coordinates correspond to the amount of any
overlap between the left and right input `Ranges`. It also returns any
metadata in the subject range if the subject overlaps the query.

@@

The `join_overlap_inner()` function will return the `Ranges` in the query that
overlap any `Ranges` in the subject. Like the `join_overlap_intersect()`
function metadata of the subject `Range` is returned if it overlaps the query.

@@

Compared with `filter_by_overlaps()` above, the overlap left join expands the
`Ranges` to give information about each interval on the query `Ranges` that
overlap those on the subject `Ranges` as well as the intervals on the left that
do not overlap any range on the right.

#### Finding your neighbours
We also provide methods for finding nearest, preceding or following `Ranges`.
Conceptually this is identical to our approach for finding overlaps, except the
semantics of the join are different.

```{r neigbours, echo = FALSE, out.width="600px"}
knitr::include_graphics(here::here("img", "neighbours.png"))
```

```{r}
join_nearest(ir0, ir1)
join_follow(ir0, ir1)
join_precede(ir0, ir1) # nothing precedes returns empty `Ranges`
join_precede(ir1, ir0)
```

### Example: dealing with multi-mapping
This example is taken from the Bioconductor support
[site](https://support.bioconductor.org/p/100046/).

We have two `Ranges` objects. The first contains single nucleotide positions
corresponding to an intensity measurement such as a ChIP-seq experiment, while
the other contains coordinates for two genes of interest.

We want to identify which positions in the `intensities` `Ranges` overlap the
genes, where each row corresponds to a position that overlaps a single gene.

First we create the two `Ranges` objects
```{r ex1}
intensities <- data.frame(seqnames = "VI",
                          start = c(3320:3321,3330:3331,3341:3342),
                          width = 1) %>%
  as_granges()

@@

### Grouping by overlaps

@@

### Data Import/Output

We provide convenience functions via **rtracklayer** and **GenomicAlignments** for
reading/writing the following data formats to/from `Ranges` objects.

@@

# Learning more

>>>>

```{r detach, include = FALSE}
detach("package:dplyr", unload = TRUE)
detach("package:plyranges", unload = TRUE)

@@

# Getting started with the **plyranges** package

## _Ranges_ revisited

In Bioconductor there are two classes, `IRanges` and `GRanges`, that are
standard data structures for representing genomics data.  Throughout this
document I refer to either of these classes as _Ranges_ if an operation can be
performed on either class, otherwise I explicitly mention if a function is
appropriate for an _IRanges_ or _GRanges_.

_Ranges_ objects can either represent sets of integers as _IRanges_ (which have
start, end and width attributes) or represent genomic intervals (which have
additional attributes, sequence name, and strand) as _GRanges_. In addition,
both types of _Ranges_ can store information about their intervals as metadata
columns (for example GC content over a genomic interval).

_Ranges_ objects follow the tidy data principle: each row of a _Ranges_ object
corresponds to an interval, while each column will represent a variable about
that interval, and generally each object will represent a single unit of
observation (like gene annotations).

Consequently, _Ranges_ objects provide a powerful representation for reasoning
about genomic data. In this vignette, you will learn more about _Ranges_
objects and how via grouping, restriction and aggregation you can perform
common data tasks.

## Constructing _Ranges_

To construct an _IRanges_ we require that there are at least two columns that
represent at either a starting coordinate, finishing coordinate or the width of
the interval.

@@

To construct a _GRanges_ we require a column that represents that sequence name
( contig or chromosome id), and an optional column to represent the
strandedness of an interval.

@@

## Arithmetic on Ranges

@@

```{r anchor_fig, echo = FALSE}
knitr::include_graphics(here::here("img", "anchors.png"), dpi = 150)
```

@@

_Ranges_ can be shifted left or right. If strand information is available we
can also shift upstream or downstream.

@@

## Grouping _Ranges_

**plyranges** introduces a new class of _Ranges_ called `RangesGrouped`, this is
a similar idea to the grouped `data.frame\tibble` in **dplyr**.

Grouping can act on either the core components or the metadata columns of a
_Ranges_ object.

@@

## Restricting _Ranges_

The verb `filter` can be used to restrict rows in the _Ranges_. Note that
grouping will cause the `filter` to act within each group of the data.
```{r}
grng %>% filter(gc < 0.3)
# filtering by group
grng_by_strand %>% filter(gc == max(gc))
```

We also provide the convenience methods `filter_by_overlaps` and
`filter_by_non_overlaps` for restricting by any overlapping _Ranges_.

@@

## Summarising _Ranges_

The `summarise` function will return a `DataFrame` because the information
required to return a _Ranges_ object is lost. It is often most useful to use
`summarise()` in combination with the `group_by()` family of functions.

@@

## Joins, or another way at looking at overlaps between _Ranges_

@@

```{r olaps-diagram, echo = FALSE}
knitr::include_graphics(here::here("img", "olap-fig.png"))
```

The first function, `join_overlap_intersect()` will return a _Ranges_ object
where the start, end, and width coordinates correspond to the amount of any
overlap between the left and right input _Ranges_. It also returns any
metadata in the subject range if the subject overlaps the query.

@@

The `join_overlap_inner()` function will return the _Ranges_ in the query that
overlap any _Ranges_ in the subject. Like the `join_overlap_intersect()`
function metadata of the subject `Range` is returned if it overlaps the query.

@@

Compared with `filter_by_overlaps()` above, the overlap left join expands the
_Ranges_ to give information about each interval on the query _Ranges_ that
overlap those on the subject _Ranges_ as well as the intervals on the left that
do not overlap any range on the right.

#### Finding your neighbours
We also provide methods for finding nearest, preceding or following _Ranges_.
Conceptually this is identical to our approach for finding overlaps, except the
semantics of the join are different.

```{r neigbours, echo = FALSE}
knitr::include_graphics(here::here("img", "neighbours.png"))
```

```{r}
join_nearest(ir0, ir1)
join_follow(ir0, ir1)
join_precede(ir0, ir1) # nothing precedes returns empty _Ranges_
join_precede(ir1, ir0)
```

#### Example: dealing with multi-mapping
This example is taken from the Bioconductor support
[site](https://support.bioconductor.org/p/100046/).

We have two _Ranges_ objects. The first contains single nucleotide positions
corresponding to an intensity measurement such as a ChIP-seq experiment, while
the other contains coordinates for two genes of interest.

We want to identify which positions in the `intensities` _Ranges_ overlap the
genes, where each row corresponds to a position that overlaps a single gene.

First we create the two _Ranges_ objects
```{r ex1}
intensities <- data.frame(seqnames = "VI",
                          start = c(3320:3321,3330:3331,3341:3342),
                          width = 1) %>%
  as_granges()

@@

## Grouping by overlaps

@@

## Reading Genomic Files

We provide convenience functions via **rtracklayer** and **GenomicAlignments** for
reading/writing the following data formats to/from _Ranges_ objects.

@@

## Learning more
